\chapter{Background and Related Work}
\label{ch:background}

This chapter places the dissertation among four bodies of work: promptable segmentation and
tracking (the model we study), the tracking metric and its detection/association decomposition (how
we score it), label-free prediction of out-of-distribution performance (the statistical idea we
extend), and distribution shift in camera-trap vision (the application and the closest empirical
neighbours). It closes by stating precisely the gap this work fills.

\section{Promptable segmentation and tracking}
\label{sec:bg-promptable}
The \emph{Segment Anything} line moved segmentation from a fixed label set to an interactive, promptable
formulation, and its successors extended prompting from points and boxes to \emph{concepts}. SAM\,3~\cite{sam3}
is a promptable model that, given a short text prompt such as \texttt{"impala"}, detects, segments, and
tracks every matching instance across a video with no training on that category --- a \emph{zero-shot},
open-vocabulary detect--segment--track capability. It is exactly this capability that conservation teams
now deploy on new species and regions, and it is the (frozen) model whose transfer we predict. SAM\,3 was
released alongside SA-FARI~\cite{safari}, a large multi-animal camera-trap tracking dataset that we use as
the substrate for our natural experiment (Section~\ref{sec:bg-safari}). We treat SAM\,3 strictly as a
fixed black box: we never fine-tune it, and the only quantity we fit is a small regression \emph{about}
its behaviour.

\section{The tracking metric and its decomposition}
\label{sec:bg-hota}
Higher Order Tracking Accuracy (HOTA)~\cite{hota} evaluates multi-object tracking by combining, over a
range of localisation thresholds, a \emph{detection} term (\texttt{DetA}: were the right objects found?)
and an \emph{association} term (\texttt{AssA}: were identities kept consistent over time?). SA-FARI scores
promptable tracking with a promptable variant, \texttt{pHOTA}, that decomposes identically into
\texttt{pDetA} and \texttt{pAssA}. This decomposition is the intellectual core of the dissertation: rather
than predict a single scalar, we ask whether \emph{finding} and \emph{following} an animal transfer for
\emph{different reasons} --- the former with how novel the animal is, the latter with how difficult the
scene is --- and we therefore model the two components separately throughout.

\section{Predictable out-of-distribution generalisation}
\label{sec:bg-predictable}
A now-mature literature shows that a model's performance under distribution shift can be anticipated
\emph{without} target labels. \citet{aotl} observe that out-of-distribution (OOD) accuracy is strongly
linearly correlated with in-distribution accuracy across many models and shifts, and \citet{agl} show that
the \emph{agreement} between a pair of models tracks OOD accuracy closely enough to estimate it from
unlabelled data alone. A parallel strand estimates accuracy from a model's own statistics on the unlabelled
target: average thresholded confidence~\cite{atc} calibrates a confidence threshold on source data and
reads off target accuracy as the mass above it, while \citet{autoeval} regress classifier accuracy directly
on a \emph{distribution-shift feature} --- the Fr\'echet distance between train and test feature sets ---
which is the closest methodological cousin to our own ``regress the score on a distance'' approach.

Two properties characterise essentially all of this work, and both are limitations we depart from. First,
it targets \textbf{image classifiers} with a single scalar accuracy. The idea has only recently reached
structured outputs: \citet{pcr} estimate object-detection performance without ground truth from the spatial
consistency and confidence reliability of pre-NMS boxes, and reverse classification accuracy~\cite{rca}
predicts per-image segmentation quality by training a reverse model against a labelled reference. These are
the closest structured-output predecessors, but both operate on \emph{static single images}, neither concerns
identity over time, and none decomposes the estimate into detection versus association. Second, every one of
these methods still \emph{runs the model on the target} and inspects its predictions, confidences, or
agreement --- it estimates the score of an \emph{already-executed} model. We instead forecast from properties
knowable \emph{before the model is ever run} on the target.

\section{Distribution shift in camera-trap and wildlife vision}
\label{sec:bg-cameratrap}
That camera-trap models degrade at new locations is well established. \emph{Terra Incognita}~\cite{terra}
showed that recognisers excellent at trained camera sites fail sharply at unseen ones; WILDS~\cite{wilds}
consolidated this (via iWildCam) as a canonical real-world shift; and PanAf-FGBG~\cite{panaf} demonstrated
that behaviour-correlated \emph{backgrounds} causally drive the drop, proposing a latent-space mitigation.
These works are the empirical backdrop for our environment axis --- but they share two features that mark the
gap. They \textbf{measure or mitigate} degradation after the fact rather than \emph{predicting} per-site or
per-species reliability in advance, and they study \textbf{classifiers or behaviour recognisers}, not
promptable trackers, scored by a single accuracy with no detection/association split. Most tellingly, a
recent unified study of camera-trap recognition over time~\cite{jeon} explicitly names ``predicting when
zero-shot models will succeed at a new site'' as an open question --- and stops at naming it. This
dissertation builds the predictor that question asks for.

\section{What should predict transfer}
\label{sec:bg-predictors}
Our candidate predictors are motivated by representation-learning results. Vision foundation models encode
taxonomic structure across the tree of life~\cite{bioclip}, which motivates a \emph{taxonomic} distance and,
because appearance and phylogeny are correlated, a \emph{visual} distance; we compute the latter from
mask-cropped DINOv2~\cite{dinov2} embeddings, with CLIP~\cite{clip} as a robustness variant. The camera-trap
location results above~\cite{terra,panaf} motivate an \emph{environment} distance from the scene background,
and per-video timestamps motivate a \emph{temporal} gap. Crucially, all four are computable without any label
of the target animal, and all are measured against a frozen reference set, so no target information can leak
into a predictor.

\section{The SA-FARI benchmark}
\label{sec:bg-safari}
SA-FARI~\cite{safari} is a large open multi-animal camera-trap tracking dataset (thousands of densely
annotated videos across roughly $80$ annotated species with a seven-level taxonomy, hundreds of locations on
four continents, and per-video timestamps), released with SAM\,3 and scored by the official \texttt{VEval}
evaluator. It ships with abundant \emph{hard negatives} --- prompts for absent species that must return
nothing --- which we keep, as they drive the false-positive side of detection. Two properties of the shipped
split shape our design and merit a precise statement. Inspection of the released annotations shows the
train/test split is disjoint by \textbf{location} (test camera sites are unseen) but that its \emph{species
are largely shared} across the split; a genuine ``unseen species'' axis is therefore not directly available
from the official split. Because SAM\,3 is frozen and zero-shot on \emph{all} of SA-FARI, however, the split
reflects nothing the model learned --- it is only our choice of a \emph{reference distribution} against which
distances are measured. This licenses a species-disjoint \emph{reference/probe partition} of the same videos
for the primary (species-novelty) experiment, complemented by the official location split for the environment
experiment; the two are near-orthogonal probes of the same data, and their construction is detailed in
Chapter~\ref{ch:methodology}.

\section{Summary and gap}
\label{sec:bg-gap}
Predictable OOD generalisation is established for classifiers scored by a scalar~\cite{aotl,agl,atc,autoeval}
and, very recently, for static detectors and segmenters~\cite{pcr,rca}; camera-trap vision has thoroughly
documented, but not forecast, location-driven degradation~\cite{terra,wilds,panaf,jeon}. No prior work
predicts, \emph{before running the model} and from label-free distances, the zero-shot transfer of a
\textbf{promptable video tracker}, and none separates that prediction into \textbf{detection} and
\textbf{association}. Asking that question rigorously --- an out-of-sample test, over SA-FARI's
species-\allowbreak by-\allowbreak place-\allowbreak by-\allowbreak time structure, of whether label-free distances
predict SAM\,3's transfer, decomposed into \texttt{pDetA} and \texttt{pAssA} --- is the contribution of this
dissertation. As Chapter~\ref{ch:results} reports, the honest answer is negative: transfer is not
distance-driven.
