\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary of findings}
% TODO: expand.
We asked whether SAM\,3's zero-shot promptable-tracking accuracy on an unseen (species, place) can be
predicted, before running the model and without target labels, from four label-free distances --- separately
for detection and association. Under a deliberately powerful out-of-sample test the answer is no: no distance
beats a mean-predictor baseline on either a species or a location hold-out, the one apparent effect is an
animal-size confound, and a follow-up intrinsic-difficulty pivot collapses to the same size effect. Only object
size predicts detection, modestly, and it is the nuisance the analysis controls for. Association carries almost
no signal distinct from detection. The result is a rigorous, decomposed null.

\section{Contributions}
% TODO: expand.
To our knowledge this is the first study to test, out of sample, whether a \emph{promptable video tracker's}
zero-shot transfer is predictable from label-free, before-running signals, and the first to decompose that
question into detection and association. Its contribution is the honest negative answer and its explanation:
transfer is not distance-driven; the apparent signals reduce to an animal-size confound; and the test is shown
to detect that one real signal when present, so the null reflects an absence of signal, not of power.

\section{Future work}
% TODO: expand.
Two directions remain open. First, representational probing --- whether the target species is separable in
SAM\,3's own feature space --- as a hedge against the unknown pretraining corpus. Second, broadening beyond a
single dataset and backbone, and to targets with richer association structure (crowded, multi-object scenes)
where the detection--association decomposition would have more to distinguish.
