\chapter{Discussion}
\label{ch:discussion}

\section{Interpretation}
% TODO: expand.
Neither hypothesis is supported. H1 (detection falls with species novelty) and H2 (association falls with
environment difficulty) both fail the pre-registered bar: no label-free distance beats a mean predictor out of
sample, and the one coefficient that reached significance --- visual distance, with the wrong sign --- is an
animal-size artefact. The follow-up intrinsic-difficulty pivot fares no better: low-light, clutter and night/IR
predict detection only by proxying animal size, which alone carries the out-of-sample gain. So the honest
reading is H0: transfer is \emph{not} distance-driven, and the only label-free property that predicts detection
is object size --- the very quantity introduced as a nuisance control.

The detection--association decomposition, the intellectual core of the design, has a second, quieter finding:
association is a near-degenerate target on this data. Because most cells contain a single object,
\texttt{pAssA} tracks \texttt{pDetA} almost exactly, leaving little identity-specific variance for any feature
to predict. The decomposition is therefore honestly a detection story; we report the association null with
those divergence figures as its explanation rather than as a separate failed search.

\section{Why label-free distances do not yield a deployment estimator}
% TODO: expand.
The field problem that motivated this work --- telling a practitioner, before spending compute, whether SAM\,3
is trustworthy for a new (species, place) --- is not solved by label-free distances. That is a negative but
useful answer: the intuition that ``far-from-training implies unreliable'' does not hold for a promptable video
tracker, so a deployment gate cannot be built from taxonomic, visual, environmental or temporal distance to a
reference set alone. The one predictive label-free signal, animal size, is too coarse and too obvious to serve
as such a gate. What the negative result does establish is a rigorous baseline: any future before-running
estimator must clear the same whole-group hold-out bar that these well-motivated distances failed.

\section{Limitations and threats to validity}
% TODO: expand.
Three limitations bound the claim. First, distances are measured against the SA-FARI train split, \emph{not}
SAM\,3's true pretraining corpus (Meta's SA-Co), which is undisclosed; the null is scoped to
distance-from-SA-FARI, and a coverage-anchored proxy for the real corpus remains untested (the deferred
familiarity proxy is one such hedge). Second, the analysis is a single dataset and a single frozen backbone.
Third, the effective sample is tens of independent species/locations, not hundreds of cells --- which is why
inference is group-clustered throughout, and why aggregate correlations (e.g. the location-level night/IR
effect) are reported alongside their much weaker cell-level values to avoid the ecological-correlation fallacy.
The open branch this work does not pursue is representational probing --- whether SAM\,3's own features encode
the species.

\section{Ethical considerations}
% TODO: expand.
The data are anonymised camera-trap footage released under a non-commercial licence; location identifiers are
already anonymised in SA-FARI. The negative result carries a responsible-deployment message: because label-free
distance does \emph{not} reliably flag when SAM\,3 will fail, conservation teams should not treat
low-distance-from-training as a safety guarantee, and reliability should be checked with labelled spot-audits
rather than assumed.
