\chapter{Introduction}
\label{ch:introduction}

\section{Motivation}
% TODO: expand into ~1 page.
Promptable video trackers such as SAM\,3~\cite{sam3} segment and track any named animal in a video
with no training on that species, and conservation teams now deploy them on new species and regions
constantly. Yet no rule tells a practitioner, \emph{before running the model}, whether its output is
trustworthy for a given animal in a given place.

\section{Problem statement and research question}
% TODO: state the RQ precisely and scope it to the SA-FARI train split.
This dissertation asks whether SAM\,3's zero-shot tracking accuracy on an unseen (species, place) can
be \emph{predicted in advance} from properties measurable without any label of the target, and which
factors govern that transfer --- separately for \emph{finding} the animal and \emph{following} it.

\section{Hypotheses}
% TODO: motivate each hypothesis from the background.
\begin{description}[leftmargin=2.6em,style=nextline]
  \item[H0 (null)] the label-free distances carry no predictive information about
    \texttt{pDetA}/\texttt{pAssA} --- transfer is not distance-driven.
  \item[H1 (detection)] \texttt{pDetA} falls with species novelty (taxonomic + visual distance).
  \item[H2 (association)] \texttt{pAssA} falls with environment difficulty (scene, day/night/IR,
    clutter, camera motion).
\end{description}
Failing to reject H0 is itself an informative result --- and, as Chapter~\ref{ch:results} shows, it is
what we find: neither a novelty distance nor a follow-up \emph{intrinsic-difficulty} pivot predicts
transfer beyond a trivial animal-size effect. Representational probing (does SAM\,3 encode taxonomy?)
remains the open branch for future work.

\section{Contributions}
% TODO: finalise once results exist.
To our knowledge this is the first study to \emph{predict}, before the model is run and from label-free
distances, the zero-shot transfer of a promptable video tracker --- and the first to split that prediction
into \textbf{detection} (\texttt{pDetA}) and \textbf{association} (\texttt{pAssA}). This distinguishes it
from label-free performance prediction for classifiers, which estimates a single scalar from the model's own
outputs on the target~\cite{aotl,agl,atc}, and from the very recent static-image detector/segmenter
estimators~\cite{pcr,rca}; and from camera-trap shift studies, which measure but do not forecast per-site
reliability~\cite{terra,panaf,jeon}. Concretely we deliver: (i) a rigorous, out-of-sample \emph{negative}
result --- under a deliberately powerful test (whole-species and whole-location hold-out with
group-cluster-bootstrap significance), label-free before-running distances do \emph{not} predict SAM\,3's
transfer; (ii) the detection-versus-association decomposition, tested on two near-orthogonal splits --- a
constructed species hold-out (novelty) and the official location hold-out (environment); and (iii) a
\emph{hardened} account of \emph{why} it fails: a follow-up intrinsic-difficulty pivot collapses to the same
animal-size confound, so the only label-free property that predicts detection is object size --- the very
nuisance the analysis controls for --- and that positive control confirms the test detects signal when it
exists.

\section{Dissertation outline}
% TODO: update page/chapter cross-references as chapters are written.
Chapter~\ref{ch:background} reviews the related work; Chapter~\ref{ch:methodology} describes the data,
measurement, distance features, and modelling; Chapter~\ref{ch:results} reports the results and
out-of-sample validation; Chapter~\ref{ch:discussion} discusses them; Chapter~\ref{ch:conclusion} concludes.
