\chapter{Data and Methodology}
\label{ch:methodology}

\section{Dataset, splits, and the analysis unit}
\label{sec:meth-data}
We use SA-FARI~\cite{safari} throughout. In the released \texttt{\_ext} annotations, a probe is a
(video, text-prompt) pair; the animal's species and its seven-level taxonomy are resolved through the
prompt's \texttt{category\_id}, while per-video \texttt{location\_id} and creation timestamp give place and
time. We aggregate scores to a \emph{cell} --- a (species, location, time) group --- and attach support
counts ($n_\text{frames}$, $n_\text{masklets}$, $n_\text{videos}$) used for weighting and as covariates.
Hard negatives (prompts for absent species) are kept, as they drive the false-positive side of detection.

\paragraph{Two complementary splits.}
Inspection of the annotations shows the shipped split is disjoint by \emph{location} but shares species
across train and test (Section~\ref{sec:bg-safari}); on that split species-distance is $\approx 0$ for every
cell, so it cannot test species novelty. Because SAM\,3 is frozen and zero-shot on all of SA-FARI, the split
only defines our \emph{reference distribution}, so we are free to repartition the same videos. We therefore
use two near-orthogonal experiments and run inference only once (scoring is per-video and split-agnostic):
\begin{description}[leftmargin=1.6em,style=nextline]
  \item[Split A --- species hold-out (primary).] The pooled species are partitioned into a reference set and
    a held-out set, stratified by taxonomic group so held-out species span a range of novelty. Species-distance
    varies while the environment stays largely familiar (held-out species recur at reference locations),
    isolating H1. Distances use a \emph{leave-species-out} reference and it is validated by leave-species-out
    cross-validation.
  \item[Split B --- location hold-out (secondary).] The official train/test split, whose unseen locations
    isolate H2 and whose benchmark comparability carries the Phase-1 score-sanity check.
\end{description}
Split A is a \emph{constructed} hold-out --- legitimate because the model is frozen, and reported as such;
we cross-check aggregate scores against Split B, which is benchmark-comparable.

\section{Measuring transfer}
\label{sec:meth-measure}
SAM\,3~\cite{sam3} is run frozen over the probe videos (species-specific prompts as the primary condition,
a generic ``animal'' prompt for robustness) and scored with the official \texttt{VEval} evaluator --- the
metric is never re-implemented --- giving per-cell \texttt{pDetA}/\texttt{pAssA}/\texttt{pHOTA}~\cite{hota}
with their support. A single inference pass over the union of Split~A and Split~B probes serves both
experiments.

\section{The frozen reference (leakage firewall)}
\label{sec:meth-reference}
For each split the reference (``seen'') species, locations, and taxonomy are frozen once; every distance is
then computed against that reference only, so no probe information can enter a feature --- verified by a unit
test that swaps the reference and checks the distances move as expected. We assert the axis that is genuinely
disjoint (location for Split~B; the held-out species for Split~A) and \emph{report} any residual species
overlap honestly rather than asserting it away. On Split~A, distances use a leave-species-out view of the
reference, so a held-out species' distance is to the nearest \emph{other} reference species.

\section{Label-free distance features}
\label{sec:meth-features}
All four distances (plus a proxy) are label-free for the target and computed against the frozen reference.
\subsection{Taxonomic distance}
Lowest-common-ancestor tree steps from the probe's species to the nearest reference species (shared genus
$=1$, family $=2$, order $=3$, \dots); the leave-species-out reference keeps this non-trivial for shared
species.
\subsection{Visual distance}
Cosine distance between DINOv2~\cite{dinov2} embeddings of the \emph{ground-truth-mask-cropped} animal and
the nearest reference-species prototype, with a Fr\'echet/MMD variant and a CLIP~\cite{clip} encoder as
robustness checks. Using the dataset's ground-truth masks means this feature needs no SAM\,3.
\subsection{Environment distance}
Embedding distance between the animal-masked-out background scene and the nearest reference location, plus an
interpretable day/night--infrared flag derived from colour statistics.
\subsection{Temporal gap}
The minimum absolute year gap between a cell's footage and the nearest reference footage.
\subsection{SAM\,3 familiarity proxy}
Separability of the target species in SAM\,3's own feature space --- a hedge against the unknown true
pretraining set, reported alongside the other distances rather than treated as ground truth.

\section{Modelling the law}
\label{sec:meth-model}
Each bounded target (\texttt{pDetA}, \texttt{pAssA}) is modelled \emph{separately} with a support-weighted
beta / logit-link regression that includes a $\log(n_\text{frames})$ covariate, so that ``rare'' is never
mistaken for ``far''. Variance is attributed to each factor by dominance/commonality analysis with a VIF
collinearity check --- not by four univariate fits --- and the headline contrast (does detection load on
species novelty and association on environment?) is quantified across the two targets. The fitted models are
validated \emph{out of sample} by holding out whole groups: leave-species-out on Split~A and
leave-location-out on Split~B, with bootstrap confidence intervals reported wherever a point estimate
appears. The validated models are wrapped into the reliability estimator: four distances in, predicted
\texttt{pDetA}/\texttt{pAssA} with an interval out.

\section{Reproducibility and constraints}
\label{sec:meth-repro}
The model is frozen and scored by the official evaluator; the seen set is frozen before any distance is
computed; hard negatives are retained; and all inference is single-GPU and inference-only. Every reported
number is regenerated from the repository, and the split definitions, seeds, and reference sets are persisted
so both experiments are reproducible end to end.
