% =====================================================================
%  MSc dissertation — Predicting the Zero-Shot Transfer of a Promptable
%  Video Tracker (SA-FARI / SAM 3).  Skeleton / initial structure.
%  Build:  report/build.sh report/dissertation   (Tectonic + biber + gs)
% =====================================================================
\documentclass[11pt,a4paper]{report}

\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \IfFileExists{lmodern.sty}{\usepackage{lmodern}\usepackage{microtype}}{}
\else
  \usepackage{fontspec}
  \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\fi
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[section]{placeins} % keep each table/float inside the section that references it
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{enumitem}
\usepackage{caption}
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,calc}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[colorlinks=true,linkcolor=NavyBlue,citecolor=NavyBlue,urlcolor=NavyBlue]{hyperref}

\setlist{nosep,leftmargin=1.4em}
\captionsetup{font=small,labelfont=bf}

\begin{document}

% -------------------------------------------------------------- title page
\begin{titlepage}
\centering
\vspace*{2cm}
{\LARGE\bfseries Predicting the Zero-Shot Transfer of a\\[4pt]
Promptable Video Tracker\par}
\vspace{6pt}
{\large Forecasting SAM\,3's per-species, per-place reliability ---\\
for detection and association --- from label-free distances\par}
\vspace{2cm}
{\large\bfseries Kostiantyn Boiar\par}
{\normalsize Student ID: \texttt{<TODO>}\par}   % TODO: student number
\vspace{1.5cm}
{\normalsize Submitted in partial fulfilment of the requirements for the\\
Degree of Master of Science in \textbf{<TODO: programme>}\par}  % TODO: exact MSc programme
\vspace{0.8cm}
{\normalsize School of Computing Science\\
University of Glasgow\par}
\vspace{1.5cm}
{\normalsize Supervisor: Dr Marwa Mahmoud\par}
\vspace{0.6cm}
{\normalsize \today\par}
\vfill
\end{titlepage}

\pagenumbering{roman}

% -------------------------------------------------------------- declaration
\chapter*{Education Use Consent \& Declaration}
% TODO: replace with the School of Computing Science's official wording.
\noindent\textbf{Education Use Consent.}\\
I hereby give my permission for this project to be shown to other University of Glasgow students and
to be distributed in an electronic format.\\[10pt]
\begin{tabular}{@{}p{3cm}p{5cm}p{2cm}p{5cm}@{}}
Name: & \hrulefill & Signature: & \hrulefill
\end{tabular}
\vspace{1.5cm}

\noindent\textbf{Declaration of Originality.}\\
I declare that this dissertation is my own work, except where explicitly stated and referenced, and
that it has not been submitted for any other degree at the University of Glasgow or any other
institution.

% -------------------------------------------------------------- abstract
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
Promptable trackers such as SAM\,3 segment and track any named animal in a video with no training on
that species, yet no rule tells a practitioner \emph{in advance} whether the model is reliable for a
given animal in a given place. This dissertation asks whether such a rule exists. On the SA-FARI
benchmark, whose splits are disjoint by species and location, we measure SAM\,3's zero-shot
promptable-tracking accuracy with the official evaluator, decompose it into detection (\texttt{pDetA})
and association (\texttt{pAssA}), and test whether each is predicted, \emph{before running the model}
and without any target label, by four distances (taxonomic, visual, environment, temporal). Under a
deliberately powerful test --- whole-species and whole-location hold-out with group-cluster-bootstrap
significance --- \textbf{it is not}: no distance beats a mean-predictor baseline out of sample, and the
one apparent effect is an animal-size confound. A follow-up pivot to intrinsic scene difficulty
(low-light, clutter) collapses to the same size effect, and the decomposition shows association carries
almost no signal distinct from detection. The contribution is thus a rigorous, decomposed \emph{negative}
result --- the first for a promptable video tracker --- validated by a test shown to detect the one real
signal (object size) when it is present.

% -------------------------------------------------------------- acknowledgements
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
% TODO: thank supervisor, dataset providers, etc.
I thank my supervisor, Dr Marwa Mahmoud, for guidance throughout this project.

% -------------------------------------------------------------- contents
\tableofcontents
\listoffigures
\listoftables

% -------------------------------------------------------------- body
\clearpage
\pagenumbering{arabic}

\input{chapters/introduction}
\input{chapters/background}
\input{chapters/methodology}
\input{chapters/results}
\input{chapters/discussion}
\input{chapters/conclusion}

% -------------------------------------------------------------- references
\clearpage
\addcontentsline{toc}{chapter}{Bibliography}
\bibliographystyle{plainnat}
\bibliography{references}

% -------------------------------------------------------------- appendices
\appendix
\input{chapters/appendices}

\end{document}
